---
title: "ANALYSIS OF IMPACT OF WEATHER ON ROAD ACCIDENTS"
author: "HITHESRAJA NAVARETHINAM, LEEKSHITHA RESOJU, LOKESH REDDY KONREDDY"
date: "2025-04-12"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Motivation and Overview
Road accidents are a significant global issue, responsible for over 1.3 million deaths each year. While human error is a well-studied cause, adverse weather conditions—like rain, fog, and snow—also play a crucial role by impairing visibility and reducing road traction. This project explores how different weather conditions influence traffic accident patterns in California. Our goal was to uncover data-driven insights that can support smarter traffic planning and road safety measures.
## Data
We used two key datasets from Kaggle:
Accident Data: Includes start/end times, location (county), accident duration, and severity.
Weather Data: Contains temperature, visibility, precipitation, and categorical weather types.
After cleaning and formatting, we joined the datasets by date and county to create a reliable dataset for further analysis. Additional features like time of day, season, and weekend indicators were engineered to enrich our analysis.

## Exploratory Data Analysis
Our EDA uncovered clear patterns. Accident counts peaked around 4–5 AM and were higher on Fridays. Winter had the most accidents, especially under rain and fog. Weekends showed fewer accidents but longer durations. Seasonal and time-of-day visualizations revealed clear risk trends, supporting the development of more targeted traffic safety policies.

## Data Analysis
We applied several statistical and machine learning methods:
ANOVA and Chi-square tests validated relationships between season, time, and accident duration.
Logistic Regression predicted if an accident would exceed 60 minutes, with ~85% accuracy.
Random Forest Regression offered precise duration predictions (R² = 0.71).
Clustering grouped weather types into high-risk (rain, fog), low-risk (clear), and long-duration clusters (snow).
The results emphasized the importance of seasonality, time of day, and weather in shaping accident risks.

## Narrative Summary:
This project bridges the gap between environmental conditions and road safety outcomes. We discovered that adverse weather significantly influences both the frequency and severity of accidents. Winter is the most dangerous season, rain and fog are top risk factors, and evenings are particularly vulnerable periods. Our machine learning models showed strong predictive power, and our Shiny app empowers users to explore these trends interactively. Together, these tools provide actionable insights for public safety improvements and smarter traffic planning. 

## Git HUb Link : https://github.com/leekshithar/Final-Project

## Weather Impact on California Traffic Accidents Analysis
```{r}
library(dplyr); library(ggplot2); library(lubridate); library(randomForest); library(caret)
library(gridExtra); library(scales); library(tidyr); library(cluster); library(ggdendro)
set.seed(123)

# 1. DATA LOADING AND PREPROCESSING
accidents_data <- read.csv("C:/Users/hithe/Desktop/project R/Final accident data set.csv", stringsAsFactors = FALSE)
weather_data <- read.csv("C:/Users/hithe/Desktop/project R/Final weather data set.csv", stringsAsFactors = FALSE)
cat("Accident data dimensions:", nrow(accidents_data), "x", ncol(accidents_data))
cat("\nWeather data dimensions:", nrow(weather_data), "x", ncol(weather_data), "\n")

# Process date-time fields
process_datetime <- function(data, start_col, end_col) {
  data$start_time <- as.POSIXct(data[[start_col]], format="%m/%d/%Y %H:%M", tz="UTC")
  if(all(is.na(data$start_time))) {
    formats <- c("%Y-%m-%d %H:%M:%S", "%m/%d/%Y %H:%M:%S", "%d-%m-%Y %H:%M")
    for(fmt in formats) {
      test_convert <- as.POSIXct(data[[start_col]][1:10], format=fmt, tz="UTC")
      if(!all(is.na(test_convert))) {
        data$start_time <- as.POSIXct(data[[start_col]], format=fmt, tz="UTC")
        data$end_time <- as.POSIXct(data[[end_col]], format=fmt, tz="UTC"); break
      }
    }
  } else { data$end_time <- as.POSIXct(data[[end_col]], format="%m/%d/%Y %H:%M", tz="UTC") }
  
  data$date <- as.Date(data$start_time)
  data$hour <- hour(data$start_time)
  data$day_of_week <- wday(data$start_time, label=TRUE)
  data$month_num <- month(data$start_time)
  data$month <- month(data$start_time, label=TRUE)
  data$year <- year(data$start_time)
  data$time_of_day <- case_when(
    data$hour >= 5 & data$hour < 12 ~ "Morning",
    data$hour >= 12 & data$hour < 17 ~ "Afternoon",
    data$hour >= 17 & data$hour < 21 ~ "Evening",
    TRUE ~ "Night"
  )
  if(!any(is.na(data$start_time)) && !any(is.na(data$end_time))) {
    data$duration_mins <- as.numeric(difftime(data$end_time, data$start_time, units="mins"))
    data$duration_mins[data$duration_mins < 0 | data$duration_mins > 1440] <- NA
  }
  return(data)
}

accidents_data <- process_datetime(accidents_data, "StartTime.UTC.", "EndTime.UTC.")
weather_data <- process_datetime(weather_data, "StartTime.UTC.", "EndTime.UTC.")

# Create season variable
accidents_data$season <- case_when(
  accidents_data$month_num %in% c(12, 1, 2) ~ "Winter",
  accidents_data$month_num %in% c(3, 4, 5) ~ "Spring",
  accidents_data$month_num %in% c(6, 7, 8) ~ "Summer",
  accidents_data$month_num %in% c(9, 10, 11) ~ "Fall",
  TRUE ~ NA_character_
)
weather_data$season <- case_when(
  weather_data$month_num %in% c(12, 1, 2) ~ "Winter",
  weather_data$month_num %in% c(3, 4, 5) ~ "Spring",
  weather_data$month_num %in% c(6, 7, 8) ~ "Summer",
  weather_data$month_num %in% c(9, 10, 11) ~ "Fall",
  TRUE ~ NA_character_
)

# Create county standardized field for joining
accidents_data$county_std <- tolower(gsub("[^a-zA-Z0-9]", "", accidents_data$County))
weather_data$county_std <- tolower(gsub("[^a-zA-Z0-9]", "", weather_data$County))

# 2. BASIC VISUALIZATIONS
p1 <- ggplot(accidents_data, aes(x = hour)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Accidents by Hour of Day", x = "Hour", y = "Number of Accidents") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(accidents_data, aes(x = day_of_week)) +
  geom_bar(fill = "darkred", color = "black") +
  labs(title = "Accidents by Day of Week", x = "Day of Week", y = "Number of Accidents") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

grid.arrange(p1, p2, ncol = 2)

# 3. SEASONAL ANALYSIS
if(any(!is.na(accidents_data$season))) {
  seasonal_accidents <- accidents_data %>%
    filter(!is.na(season)) %>%
    group_by(season) %>%
    summarize(
      accident_count = n(),
      accidents_per_day = n() / n_distinct(date),
      avg_duration = mean(duration_mins, na.rm = TRUE),
      .groups = 'drop'
    )
  season_levels <- c("Winter", "Spring", "Summer", "Fall")
  seasonal_accidents$season <- factor(seasonal_accidents$season, levels = season_levels)
  
  seasonal_weather <- weather_data %>%
    filter(!is.na(season)) %>%
    group_by(season, Type) %>%
    summarize(count = n(), .groups = 'drop') %>%
    group_by(season) %>%
    mutate(percent = count / sum(count) * 100) %>%
    ungroup() %>%
    mutate(season = factor(season, levels = season_levels))
  
  p5 <- ggplot(seasonal_accidents, aes(x = season, y = accident_count, fill = season)) +
    geom_bar(stat = "identity", color = "black") +
    labs(title = "Accident Count by Season", x = "Season", y = "Number of Accidents") +
    theme_minimal() + theme(plot.title = element_text(face = "bold"), legend.position = "none")
  
  p6 <- ggplot(seasonal_accidents, aes(x = season, y = avg_duration, fill = season)) +
    geom_bar(stat = "identity", color = "black") +
    labs(title = "Average Accident Duration by Season", x = "Season", y = "Duration (minutes)") +
    theme_minimal() + theme(plot.title = element_text(face = "bold"), legend.position = "none")
  
  p7 <- ggplot(seasonal_weather, aes(x = season, y = count, fill = Type)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Weather Types by Season", x = "Season", y = "Number of Weather Events") +
    theme_minimal() + theme(plot.title = element_text(face = "bold"))
  
  grid.arrange(p5, p6, p7, layout_matrix = rbind(c(1,2), c(3,3)))
}

# 4. MONTHLY ANALYSIS (if seasonal not available)
if(!exists("seasonal_accidents") || nrow(seasonal_accidents) == 0) {
  monthly_accidents <- accidents_data %>%
    group_by(month) %>%
    summarize(
      accident_count = n(),
      avg_duration = mean(duration_mins, na.rm = TRUE),
      .groups = 'drop'
    )
  month_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  monthly_accidents$month <- factor(monthly_accidents$month, levels = month_order)
  
  p5a <- ggplot(monthly_accidents, aes(x = month, y = accident_count, fill = month)) +
    geom_bar(stat = "identity", color = "black") +
    labs(title = "Accident Count by Month", x = "Month", y = "Number of Accidents") +
    theme_minimal() + theme(plot.title = element_text(face = "bold"), 
                            legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
  
  p6a <- ggplot(monthly_accidents, aes(x = month, y = avg_duration, fill = month)) +
    geom_bar(stat = "identity", color = "black") +
    labs(title = "Average Accident Duration by Month", x = "Month", y = "Duration (minutes)") +
    theme_minimal() + theme(plot.title = element_text(face = "bold"), 
                            legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
  
  if(n_distinct(weather_data$Type) > 1) {
    monthly_weather <- weather_data %>%
      group_by(month, Type) %>%
      summarize(count = n(), .groups = 'drop') %>%
      mutate(month = factor(month, levels = month_order))
    
    p7a <- ggplot(monthly_weather, aes(x = month, y = count, fill = Type)) +
      geom_bar(stat = "identity", position = "stack") +
      labs(title = "Weather Types by Month", x = "Month", y = "Number of Weather Events") +
      theme_minimal() + theme(plot.title = element_text(face = "bold"),
                              axis.text.x = element_text(angle = 45, hjust = 1))
    
    grid.arrange(p5a, p6a, p7a, layout_matrix = rbind(c(1,2), c(3,3)))
  } else {
    grid.arrange(p5a, p6a, ncol = 2)
  }
}

# 5. TIME OF DAY & REGIONAL ANALYSIS
tod_accidents <- accidents_data %>%
  group_by(time_of_day) %>%
  summarize(
    accident_count = n(),
    avg_duration = mean(duration_mins, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")))

p9 <- ggplot(tod_accidents, aes(x = time_of_day, y = accident_count, fill = time_of_day)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Accidents by Time of Day", x = "Time of Day", y = "Number of Accidents") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"), legend.position = "none")

if(n_distinct(weather_data$Type) > 1) {
  tod_weather <- weather_data %>%
    group_by(time_of_day, Type) %>%
    summarize(count = n(), .groups = 'drop') %>%
    mutate(time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")))
  
  p10 <- ggplot(tod_weather, aes(x = time_of_day, y = count, fill = Type)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = "Weather Types by Time of Day", x = "Time of Day", y = "Number of Weather Events") +
    theme_minimal() + theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
  
  grid.arrange(p9, p10, ncol = 2)
} else {
  print(p9)
}

county_accidents <- accidents_data %>%
  group_by(County) %>%
  summarize(
    accident_count = n(),
    avg_duration = mean(duration_mins, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(accident_count))
top_counties <- head(county_accidents, 10)

p11 <- ggplot(top_counties, aes(x = reorder(County, accident_count), y = accident_count)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  coord_flip() +
  labs(title = "Top 10 Counties by Accident Count", x = "County", y = "Number of Accidents") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

print(p11)

# 6. STATISTICAL TESTS
# Create weekend and duration category variables
accidents_data$day_num <- wday(accidents_data$start_time)
accidents_data$is_weekend <- ifelse(accidents_data$day_num %in% c(1, 7), 1, 0)
accidents_data$weekend_factor <- factor(accidents_data$is_weekend, levels = c(0, 1), 
                                        labels = c("Weekday", "Weekend"))
accidents_data$duration_category <- cut(accidents_data$duration_mins,
                                        breaks = c(0, 15, 30, 60, Inf),
                                        labels = c("Very Short", "Short", "Medium", "Long"),
                                        include.lowest = TRUE)

# ANOVA test
anova_weekend <- aov(duration_mins ~ weekend_factor, data = accidents_data)
cat("\nANOVA Test: Weekend vs Weekday Impact on Duration\n")
print(summary(anova_weekend))

# Weekend vs weekday plot
p_weekend <- ggplot(accidents_data, aes(x = weekend_factor, y = duration_mins)) +
  geom_boxplot(fill = c("skyblue", "lightgreen"), width = 0.6, 
               outlier.shape = 1, outlier.size = 2, outlier.alpha = 0.5) +
  labs(title = "Accident Duration: Weekday vs Weekend", x = "Day Type", y = "Duration (minutes)") +
  theme_minimal() + theme(plot.title = element_text(face = "bold")) +
  scale_y_continuous(limits = c(0, min(300, max(accidents_data$duration_mins, na.rm = TRUE))))

# Chi-square test
season_duration_table <- table(accidents_data$season, accidents_data$duration_category)
chi_season_duration <- chisq.test(season_duration_table)
cat("\nChi-Square Test: Season vs Duration Category\n")
print(chi_season_duration)

# Season vs duration plot
season_duration_df <- as.data.frame(season_duration_table)
names(season_duration_df) <- c("Season", "Duration", "Freq")
season_duration_df <- season_duration_df %>%
  group_by(Season) %>%
  mutate(Percentage = Freq / sum(Freq) * 100) %>%
  ungroup()

p_season <- ggplot(season_duration_df, aes(x = Season, y = Percentage, fill = Duration)) +
  geom_bar(stat = "identity", position = "fill", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Association: Season and Accident Duration", y = "Proportion") +
  theme_minimal() + theme(plot.title = element_text(face = "bold")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))

grid.arrange(p_weekend, p_season, ncol = 2)

# 7. ADVANCED MODELING
# Create a modeling dataset
model_data <- accidents_data %>%
  filter(!is.na(duration_mins), !is.na(season), !is.na(time_of_day)) %>%
  mutate(
    long_duration = ifelse(duration_mins > 60, 1, 0),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")),
    day_of_week = factor(day_of_week),
    hour_group = cut(hour, breaks = c(0, 6, 12, 18, 24), 
                     labels = c("Night (0-6)", "Morning (6-12)", 
                                "Afternoon (12-18)", "Evening (18-24)"))
  )

# Split data
set.seed(123)
train_index <- createDataPartition(model_data$long_duration, p = 0.7, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

# 7.1 LOGISTIC REGRESSION
logistic_model <- glm(
  long_duration ~ season + time_of_day + day_of_week + hour_group,
  family = binomial(link = "logit"),
  data = train_data
)

# Make predictions and evaluate
logistic_pred_prob <- predict(logistic_model, newdata = test_data, type = "response")
logistic_pred_class <- ifelse(logistic_pred_prob > 0.5, 1, 0)
logistic_conf_matrix <- table(Predicted = logistic_pred_class, Actual = test_data$long_duration)
logistic_accuracy <- sum(diag(logistic_conf_matrix)) / sum(logistic_conf_matrix)

# Visualize logistic regression coefficients
coef_data <- data.frame(
  feature = names(coef(logistic_model)[-1]),
  coefficient = coef(logistic_model)[-1],
  p_value = summary(logistic_model)$coefficients[-1, 4]
) %>%
  mutate(
    significant = p_value < 0.05,
    feature = gsub("season|time_of_day|day_of_week|hour_group", "", feature),
    feature = gsub("^", "", feature)
  ) %>%
  arrange(p_value)

p_logistic <- ggplot(head(coef_data, 10), 
                     aes(x = reorder(feature, abs(coefficient)), y = coefficient, fill = significant)) +
  geom_bar(stat = "identity") + coord_flip() +
  scale_fill_manual(values = c("gray70", "steelblue")) +
  labs(title = "Top Predictors of Long-Duration Accidents", x = "Feature", y = "Coefficient") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

print(p_logistic)

# 7.2 POISSON REGRESSION
poisson_data <- accidents_data %>%
  group_by(season, time_of_day, day_of_week) %>%
  summarize(accident_count = n(), .groups = 'drop') %>%
  filter(!is.na(season), !is.na(time_of_day), !is.na(day_of_week)) %>%
  mutate(
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")),
    day_of_week = factor(day_of_week)
  )

poisson_train_index <- createDataPartition(poisson_data$accident_count, p = 0.7, list = FALSE)
poisson_train <- poisson_data[poisson_train_index, ]
poisson_test <- poisson_data[-poisson_train_index, ]

poisson_model <- glm(
  accident_count ~ season + time_of_day + day_of_week,
  family = poisson(link = "log"),
  data = poisson_train
)

poisson_pred <- predict(poisson_model, newdata = poisson_test, type = "response")
poisson_rmse <- sqrt(mean((poisson_test$accident_count - poisson_pred)^2))

# Poisson model visualization
poisson_results <- data.frame(actual = poisson_test$accident_count, predicted = poisson_pred)
p_poisson <- ggplot(poisson_results, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Poisson Model: Actual vs Predicted Counts", x = "Actual", y = "Predicted") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

print(p_poisson)

# 7.3 RANDOM FOREST
rf_data <- model_data %>%
  select(duration_mins, season, time_of_day, day_of_week, hour, month_num, is_weekend) %>%
  mutate(across(c(season, time_of_day, day_of_week), as.factor))

rf_train_index <- createDataPartition(rf_data$duration_mins, p = 0.7, list = FALSE)
rf_train <- rf_data[rf_train_index, ]
rf_test <- rf_data[-rf_train_index, ]

rf_model <- randomForest(
  duration_mins ~ .,
  data = rf_train,
  ntree = 200,
  importance = TRUE
)

rf_predictions <- predict(rf_model, newdata = rf_test)
rf_rmse <- sqrt(mean((rf_test$duration_mins - rf_predictions)^2))
rf_r2 <- 1 - sum((rf_test$duration_mins - rf_predictions)^2) / 
  sum((rf_test$duration_mins - mean(rf_test$duration_mins))^2)

# RF visualization - importance plot and predictions
imp_data <- data.frame(
  Feature = rownames(importance(rf_model)),
  IncMSE = importance(rf_model)[, 1]
) %>% arrange(desc(IncMSE))

p_importance <- ggplot(imp_data, aes(x = reorder(Feature, IncMSE), y = IncMSE)) +
  geom_bar(stat = "identity", fill = "darkgreen") + coord_flip() +
  labs(title = "Random Forest Variable Importance", x = "Feature", y = "Importance") +
  theme_minimal() + theme(plot.title = element_text(face = "bold"))

p_rf_pred <- ggplot(data.frame(Actual = rf_test$duration_mins, Predicted = rf_predictions),
                    aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.3) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Random Forest: Actual vs Predicted Duration") +
  theme_minimal() + theme(plot.title = element_text(face = "bold")) +
  xlim(0, 300) + ylim(0, 300)

grid.arrange(p_importance, p_rf_pred, ncol = 2)

# 7.4 HIERARCHICAL CLUSTERING FOR WEATHER IMPACT ANALYSIS
cat("\n----- HIERARCHICAL CLUSTERING FOR WEATHER IMPACT -----\n")
cat("Using hierarchical clustering to identify weather-related accident patterns\n")

# Create a simplified dataset for clustering with key features
# First, create aggregated data by weather type
weather_impact <- accidents_data %>%
  # Join with simplified weather data
  left_join(
    weather_data %>% 
      select(date, county_std, Type),
    by = c("date" = "date", "county_std" = "county_std")
  ) %>%
  # Filter to only rows with weather data
  filter(!is.na(Type)) %>%
  # Group by weather type
  group_by(Type) %>%
  # Calculate summary statistics
  summarize(
    accident_count = n(),
    avg_duration = mean(duration_mins, na.rm = TRUE),
    med_duration = median(duration_mins, na.rm = TRUE),
    avg_hour = mean(hour, na.rm = TRUE),
    pct_weekend = mean(is_weekend, na.rm = TRUE) * 100,
    morning_pct = mean(time_of_day == "Morning", na.rm = TRUE) * 100,
    afternoon_pct = mean(time_of_day == "Afternoon", na.rm = TRUE) * 100,
    evening_pct = mean(time_of_day == "Evening", na.rm = TRUE) * 100,
    night_pct = mean(time_of_day == "Night", na.rm = TRUE) * 100,
    .groups = 'drop'
  ) %>%
  # Filter to only weather types with at least 5 accidents for reliability
  filter(accident_count >= 5)

# Print the weather impact summary
cat("\nWeather Impact Summary:\n")
print(weather_impact)

# Prepare data for hierarchical clustering
# Select numerical features
hc_data <- weather_impact %>%
  select(-Type, -accident_count) %>%  # Remove non-numeric columns
  scale()  # Standardize the data

rownames(hc_data) <- weather_impact$Type  # Use weather types as row names

# Calculate distance matrix
dist_matrix <- dist(hc_data, method = "euclidean")

# Perform hierarchical clustering
hc_result <- hclust(dist_matrix, method = "ward.D2")

# Plot dendrogram
plot(hc_result, main = "Hierarchical Clustering of Weather Types",
     xlab = "Weather Type", ylab = "Distance", hang = -1)
rect.hclust(hc_result, k = 4, border = "red")  # Draw boxes around 4 clusters

# Cut tree to get cluster assignments (using 4 clusters)
weather_clusters <- cutree(hc_result, k = 4)

# Add cluster assignments back to the data
weather_impact$cluster <- factor(weather_clusters)

# Cluster summary
weather_cluster_summary <- weather_impact %>%
  group_by(cluster) %>%
  summarize(
    weather_types = toString(Type),
    num_types = n_distinct(Type),
    total_accidents = sum(accident_count),
    avg_duration = weighted.mean(avg_duration, accident_count),
    avg_hour = weighted.mean(avg_hour, accident_count),
    pct_weekend = weighted.mean(pct_weekend, accident_count),
    .groups = 'drop'
  ) %>%
  arrange(cluster)

cat("\nWeather Cluster Summary:\n")
print(weather_cluster_summary)

# Create a more informative visualization using ggplot2
# Create dendrogram data
dendr <- as.dendrogram(hc_result)
dendr_data <- dendro_data(dendr, type = "rectangle")

# Create more elegant dendrogram
p_dendrogram <- ggplot(segment(dendr_data)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = label(dendr_data), 
            aes(x = x, y = y, label = label, color = factor(weather_clusters[label])),
            hjust = 0, angle = 90, size = 3) +
  labs(title = "Hierarchical Clustering of Weather Types",
       subtitle = "Colors indicate cluster membership",
       x = NULL, y = "Height") +
  scale_color_brewer(palette = "Set1", name = "Cluster") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

print(p_dendrogram)

# Create heatmap visualization of feature values by weather type
heatmap_data <- weather_impact %>%
  select(Type, cluster, avg_duration, avg_hour, pct_weekend,
         morning_pct, afternoon_pct, evening_pct, night_pct) %>%
  # Arrange by cluster for better visualization
  arrange(cluster)

# Prepare data for heatmap
heatmap_long <- heatmap_data %>%
  pivot_longer(cols = c(avg_duration, avg_hour, pct_weekend, 
                        morning_pct, afternoon_pct, evening_pct, night_pct),
               names_to = "Feature", values_to = "Value") %>%
  # Make feature names more readable
  mutate(Feature = case_when(
    Feature == "avg_duration" ~ "Avg Duration (min)",
    Feature == "avg_hour" ~ "Avg Hour of Day",
    Feature == "pct_weekend" ~ "Weekend %",
    Feature == "morning_pct" ~ "Morning %",
    Feature == "afternoon_pct" ~ "Afternoon %",
    Feature == "evening_pct" ~ "Evening %",
    Feature == "night_pct" ~ "Night %",
    TRUE ~ Feature
  ))

# Create a custom z-score to make the heatmap more interpretable
heatmap_long <- heatmap_long %>%
  group_by(Feature) %>%
  mutate(
    z_score = (Value - mean(Value)) / sd(Value)
  ) %>%
  ungroup()

# Create the heatmap
p_heatmap <- ggplot(heatmap_long, aes(x = Type, y = Feature, fill = z_score)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       name = "Z-score") +
  facet_grid(. ~ cluster, scales = "free_x", space = "free_x") +
  labs(title = "Weather Impact Patterns by Cluster",
       subtitle = "Red = higher than average, Blue = lower than average",
       x = "Weather Type", y = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.spacing = unit(0.5, "lines"),
        strip.text = element_text(face = "bold"))

print(p_heatmap)

# Create a parallel coordinate plot to visualize cluster characteristics
                  parallel_data <- weather_impact %>%
                    select(Type, cluster, avg_duration, avg_hour, pct_weekend, morning_pct, 
                           afternoon_pct, evening_pct, night_pct) %>%
                    # Scale the values for better comparison
                    mutate(across(c(avg_duration, avg_hour, pct_weekend, morning_pct, 
                                    afternoon_pct, evening_pct, night_pct), scale))
                  
                  parallel_long <- parallel_data %>%
                    pivot_longer(cols = c(avg_duration, avg_hour, pct_weekend, morning_pct, 
                                          afternoon_pct, evening_pct, night_pct),
                                 names_to = "Feature", values_to = "Value") %>%
                    mutate(Feature = factor(Feature, levels = c("avg_duration", "avg_hour", "pct_weekend", 
                                                                "morning_pct", "afternoon_pct", "evening_pct", "night_pct")))
                  
                  p_parallel <- ggplot(parallel_long, aes(x = Feature, y = Value, group = Type, color = cluster)) +
                    geom_line(alpha = 0.7) +
                    labs(title = "Parallel Coordinate Plot of Weather Clusters",
                         x = NULL, y = "Standardized Value", color = "Cluster") +
                    scale_color_brewer(palette = "Set1") +
                    theme_minimal() +
                    theme(plot.title = element_text(face = "bold"),
                          axis.text.x = element_text(angle = 45, hjust = 1))
                  
                  print(p_parallel)
                  
                  # Generate key insights for each cluster
                  weather_cluster_insights <- weather_impact %>%
                    group_by(cluster) %>%
                    summarize(
                      top_weather = Type[which.max(accident_count)],
                      top_weather_pct = max(accident_count) / sum(accident_count) * 100,
                      avg_duration = weighted.mean(avg_duration, accident_count),
                      primary_time = case_when(
                        weighted.mean(morning_pct, accident_count) > 
                          max(weighted.mean(afternoon_pct, accident_count), 
                              weighted.mean(evening_pct, accident_count),
                              weighted.mean(night_pct, accident_count)) ~ "Morning",
                        weighted.mean(afternoon_pct, accident_count) > 
                          max(weighted.mean(morning_pct, accident_count), 
                              weighted.mean(evening_pct, accident_count),
                              weighted.mean(night_pct, accident_count)) ~ "Afternoon",
                        weighted.mean(evening_pct, accident_count) > 
                          max(weighted.mean(morning_pct, accident_count), 
                              weighted.mean(afternoon_pct, accident_count),
                              weighted.mean(night_pct, accident_count)) ~ "Evening",
                        TRUE ~ "Night"
                      ),
                      weekend_bias = weighted.mean(pct_weekend, accident_count) > 29,  # 2/7 days are weekends (~29%)
                      .groups = 'drop'
                    )
                  
                  # 8. KEY FINDINGS AND RECOMMENDATIONS
                  cat("\n===== KEY FINDINGS =====\n")
                  
                  # Time patterns
                  peak_hour <- as.numeric(names(sort(table(accidents_data$hour), decreasing = TRUE)[1]))
                  cat("1. TEMPORAL PATTERNS:\n")
                  cat("   - Peak accident hour:", peak_hour, "\n")
                  cat("   - Most accidents during:", tod_accidents$time_of_day[which.max(tod_accidents$accident_count)], "\n")
                  if(exists("seasonal_accidents") && nrow(seasonal_accidents) > 0) {
                    cat("   - Season with highest accident rate:", 
                        as.character(seasonal_accidents$season[which.max(seasonal_accidents$accident_count)]), "\n")
                    cat("   - Season with longest durations:", 
                        as.character(seasonal_accidents$season[which.max(seasonal_accidents$avg_duration)]), "\n")
                  }
                  
                  # Weather patterns
                  if(n_distinct(weather_data$Type) > 1) {
                    weather_counts <- weather_data %>%
                      count(Type) %>%
                      arrange(desc(n)) %>%
                      mutate(percent = n / sum(n) * 100)
                    cat("\n2. WEATHER PATTERNS:\n")
                    cat("   - Most common weather type:", weather_counts$Type[1], 
                        "(", round(weather_counts$percent[1], 1), "%)\n")
                  }
                  
                  # Regional insights
                  cat("\n3. REGIONAL INSIGHTS:\n")
                  cat("   - County with highest accident frequency:", top_counties$County[1], "\n")
                  
                  # Model insights
                  cat("\n4. MODEL INSIGHTS:\n")
                  cat("   - Logistic regression accuracy:", round(logistic_accuracy * 100, 1), "%\n")
                  cat("   - Poisson model RMSE:", round(poisson_rmse, 2), "\n")
                  cat("   - Random Forest R²:", round(rf_r2, 3), "and RMSE:", round(rf_rmse, 2), "minutes\n")
                  cat("   - Top predictors of accident duration:", 
                      paste(head(rownames(importance(rf_model)), 3), collapse=", "), "\n")
                  
                  # Weather cluster insights
                  cat("\n5. HIERARCHICAL CLUSTERING INSIGHTS:\n")
                  for(i in 1:nrow(weather_cluster_insights)) {
                    cat(paste0("   - Weather Cluster ", i, ": Dominated by ", 
                               weather_cluster_insights$top_weather[i], 
                               " weather, with average accident duration of ", 
                               round(weather_cluster_insights$avg_duration[i], 1), 
                               " minutes, primarily during ", 
                               tolower(weather_cluster_insights$primary_time[i]), 
                               " hours",
                               ifelse(weather_cluster_insights$weekend_bias[i], 
                                      ", with weekend bias", 
                                      ", with weekday bias"),
                               ".\n"))
                  }
                  
                  # Recommendations
                  cat("\n===== RECOMMENDATIONS =====\n")
                  cat("1. TIME-BASED STRATEGIES:\n")
                  cat("   - Focus resources during peak hours (", peak_hour, ")\n", sep="")
                  
                  # Seasonal recommendations
                  if(exists("seasonal_accidents") && nrow(seasonal_accidents) > 0) {
                    high_freq_season <- as.character(seasonal_accidents$season[which.max(seasonal_accidents$accident_count)])
                    long_dur_season <- as.character(seasonal_accidents$season[which.max(seasonal_accidents$avg_duration)])
                    cat("\n2. SEASONAL STRATEGIES:\n")
                    cat("   - Implement preventive measures during", high_freq_season, "season\n")
                    if(high_freq_season != long_dur_season) {
                      cat("   - Prepare for longer clearance times during", long_dur_season, "season\n")
                    }
                  }
                  
                  # Location recommendations
                  cat("\n3. LOCATION-BASED STRATEGIES:\n")
                  cat("   - Prioritize improvements in", top_counties$County[1], "County\n")
                  if(n_distinct(weather_data$Type) > 1) {
                    cat("   - Implement special precautions during", weather_counts$Type[1], "conditions\n")
                  }
                  
                  # Weather cluster recommendations
                  cat("\n4. WEATHER CLUSTER STRATEGIES:\n")
                  # Find the cluster with the longest average duration
                  longest_duration_cluster <- weather_cluster_insights$cluster[
                    which.max(weather_cluster_insights$avg_duration)]
                  
                  cat(paste0("   - Prepare additional resources for accidents in Weather Cluster ", 
                             longest_duration_cluster, 
                             " (", weather_cluster_insights$top_weather[longest_duration_cluster], 
                             "-dominated), which have the longest average duration of ", 
                             round(max(weather_cluster_insights$avg_duration), 1), 
                             " minutes.\n"))
                  
                  # Find the cluster with the strongest weekend bias
                  if(any(weather_cluster_insights$weekend_bias)) {
                    weekend_clusters <- weather_cluster_insights$cluster[weather_cluster_insights$weekend_bias]
                    weekend_weathers <- weather_cluster_insights$top_weather[weather_cluster_insights$weekend_bias]
                    
                    cat(paste0("   - Schedule additional weekend coverage for Weather Cluster ", 
                               paste(weekend_clusters, collapse = ", "), 
                               " (", paste(weekend_weathers, collapse = ", "), 
                               "), which show a weekend bias in accident occurrences.\n"))
                  }
                  
                  # Provide time-of-day recommendations
                  time_recommendations <- weather_cluster_insights %>%
                    group_by(primary_time) %>%
                    summarize(
                      clusters = toString(cluster),
                      weather_types = toString(top_weather),
                      .groups = 'drop'
                    )
                  
                  for(i in 1:nrow(time_recommendations)) {
                    cat(paste0("   - Focus ", time_recommendations$primary_time[i], 
                               " resources on Weather Cluster ", time_recommendations$clusters[i], 
                               " (", time_recommendations$weather_types[i], 
                               ") conditions.\n"))
                  }
```